Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2019-12-19 manuscript.tex
reestructuracion

\documentclass[crop]{CSLB}%%%%where CSL is the template name

%The authors can define any packages after the \documentclass{CSL} command.
\usepackage{amsmath} %for dealing with mathematics,
\usepackage{amsthm} %for dealing with theorem environments,
\usepackage{cite} %%%for dealing with citations
\usepackage{hyperref} %for linking the cross references
\usepackage{graphics} %for dealing with figures.
\usepackage{algorithmic} %for describing algorithms
\usepackage{subfig} %for getting the subfigures e.g., "Figure 1a and 1b" etc.
\usepackage{url} %It provides better support for handling and breaking URLs.

%\usepackage{natbib}
\usepackage{boites}

%\usepackage{graphicx}
%\usepackage{float}
%\usepackage{psfig}
%\usepackage{amsfonts,amssymb}
%\usepackage{color}
%\usepackage{textcomp}
%\usepackage[T1]{fontenc}
%\usepackage{aecompl}
%\usepackage{times}
%\usepackage{txfonts}
%\usepackage[english]{babel}
%\usepackage[utf8x]{inputenc}
%\usepackage{amsmath}
%\usepackage{caption}
%\usepackage{siunitx}

\jname{Astrobiology}%

\newtheorem{theorem}{Theorem}
\newtheorem{condition}{Condition}

@@

%% Stochastic estimation of the waiting time for the first causal contact
%% with an extraterrestrial intelligence

@@

@@

A common approach to discussing the estimations of the number of
extraterrestrial intelligent civilizations (ETI) in the Galaxy has
been the use of the Drake equation \citep{gleiser_drake_2010,
prantzos_joint_2013, haqq-misra_drake_2017}.
%
The abundance of intelligent civilizations in the galaxy is a
longstanding question, which is often conceptualized as the problem of
the lack of received communication or the Fermi paradox
\citep{barlow_galactic_2012, Sotos_biotechnology_2019,
forgan_galactic_2016}.
%
There are many proposals aimed at solving this paradox, which make use
of statistical \citep{solomonides_probabilistic_2016,
vanhouten_isthere_2017, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}, analytical
interpretations of the Drake equation \citep{prantzos_joint_2013,
smith_broadcasting_2009} or more speculative proposals
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The uncertainties in the factors of the Drake equation make it less
prone to a formal application in order to define searching strategies
or to compute the actual number of ETIs, but is a key tool to organize
the discussion around the problem of the population of ETIs in the
Galaxy \citep{hinkel_interdisciplinary_2019}, which is noticeable
dominated by the fact that despite decades of effort not a single ETI,
if any, has been detected so far.
%
Some modifications to the original idea of the equation have tried to
imprint a stochastic nature, or to propose a probabilistic approach,
or to consider the temporal structure which is missing in the
equation.
%
Temporal aspects of the distribution of communicating civilizations
and their contacts have also been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011},
%The Impact of the Temporal Distribution of Communicating Civilizations on Their Detectability
%Temporal dispersion of the emergence of intelligence: an inter-arrival time analysis
%The spatiotemporal aspects of SETI
%
as well as efforts on considering the stochastic nature of the Drake equation
\citep{glade_stochastic_2011}.
%
The simulation approach has also benn considered
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017},
although with a similar problem that consists on the large number of
parameters which are either unknown or largely uncertain.
%
%New numerical determination of habitability in the Galaxy: the SETI connection
%
%Since the factors in the Drake equation are uncertain,
%
In this work, we propose to
avoid the frequentist approach of the Drake equation and to explore a
parameter space, where instead of computing a final number, we provide
a statistical distribution that gives conditional probabilities.
%
This is an exploratory analysis that aims at providing a numerical 
tool to discuss not the several theoretical problems summarized by the
Drake equation factors, 
but the different scenarios on the basis of statistical heuristics.
%
The approach proposed here shold be considered as a compromise between
the uncertainties of the frequentist approach and the detailed recipes
required on the simulation approach.
%
We then provide a numerical framework to explore via simulations a
parameter space of unknown observables in order to discuss different
scenarios and their consecuences in terms of the probability con
making contacts.
%
The only fact that can be stated with certainty is that for the number
of years SETI projects have been working we have not received any
signal, at least
within the technical posibilities and the 
conditions stablished by SETI \citep{tarter_search_2001}.
%
This arguably astonishing result has lead to the Fermi paradox, which
states an apparent contradiction between the high estimation of the
abundance of life in the Galaxy and the lack of their detection
\citep{vanhouten_isthere_2017}.
%
The lack of detection has also ispired to a number of alternative proposals
for new SETI strategies \citep{forgan_exoplanet_2017,
balbi_impact_2018, loeb_eavesdropping_2006, maccone_KLT_2010,
tarter_advancing_2009, enriquez_breakthrough_2017, loeb_relative_2016,
maccone_SETI_2011,  lingam_relative_2019, wright_theGsearch_2015,
maccone_SETI_2013, maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

\section{Methods and working hypotheses}\label{S_methods}
%{{{

Some complex stochastic processes can be efficiently modeled with the
discrete--event (DE) simulation approach.
%
Simulations are suitable tools to analyze systems that evolve with
time and involve randomness.
%
In general, simulations require less assumptions and simplifications
than theoretical approaches, and can be applied to systems where a
theoretical model can not be found.
%
A system described with the DE paradigm is characterized by a set of
actors and events, where actors interact causally through a series of
events on a time line and process these events in chronological order
\citep{ptolemaeus_system_2014, chung_simulation_2003,
ross_simulation_2012}.
%
In the case of study, the actors are the communicating civilizations
(CETIs), and the events are all meaningfull that produce a change in
the simulation variables, i.e., in the arrangement of CETIs and their
causal contacts.
%
Since it allows to explore several models easily, 
this method is well suited for the particular case of the difusion of
intelligence signals in the Galaxy.
%
Assuming some simple hypotheses, the discrete events method can be
performed taking into account a small number of variables, which
allow to analyze the variation of the results in the model parameter space.

In what follows, we describe the experimental setup chosen to estimate
the probabilities of causal contacts and several derived quantities in
terms of three independent parameters, namely, the mean time span
between the appeareance of consecutive communicating civilizations
($\tau_a$), their mean lifetime ($\tau_s$), and the maximum distance a
signal can be detected by another civilization ($D_{max}$).
%
Intuitively, the shorter $\tau_a$ and the larger $\tau_s$ and
$D_{max}$, there is a greater probability of the existence of causal
contacts between pairs of intelligent civilizations in the Galaxy
(Table \ref{T_simu_hypotheses}).
%
We also need to propose theoretical distribution functions for both the
distribution of lifespans and the distribution of the number of CETIs
per unit time \citep{maccone_evolution_2014,Sotos_biotechnology_2019}.
%
The shapes of these distributions are discussed in Sec.~\ref{SS_PDF_shape}.
%
A causal connection can be produced without the need of the two
involved civilizations being concurrent or active at the same time.
%
This is a fundamental property of the system which arises as a
consecuence of the large spatial and temporal scales, where a message
is transmited at light speed.
%
As a consecuence of their finite lifetime,
each CETI will fill a spherical shell region of the Galaxy, limited by two
surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central CETI until it reaches the maximum distance $D_{max}$.
%
Although a CETI could be old enough to transmit a message at large
distances, the limited power of the message and the dilution that runs
as the squared distance from the source, imposes a
detectability limit.
%
Thus, the volume reached causally by a CETI is initially a growing
sphere.
%
Following the end of the civilization, there is still a region which
is covered with the emited signals.
%
The trailing front, or surface of last contact (SLC) also grows from
the central CETI, with a delay with respect to the SFC equivalent to
the lifetime of the CETI, and produces a spherical shell region.
%
Any other CETI within this region will be aware of the originating CETI,
even if it has disapeared before the time of contact.
%
This region will grow if the SFC has not reached $D_{max}$, and
will shrink otherwise, until the SLC reaches $D_{max}$.
%
A that point, all signal from the central CETI is lost.
%
In the 2D simulation only the intersection of these spheres with the
plane of the galaxy is relevant, and produce the corresponding
circles for the filled spheres or annular regions for the spherical
shells.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic
representation of the region causally connected to the central CETI.
%
%
%
% Schematic representation of the growing communicating
%   sphere, over the space--time diagrams, where time is represented on
%   the horizontal direction, and space is represented in the vertical
%   direction.  In (a) the sphere is
%   growing as the surface of first contact has not reached the maximum
%   distance.  In (b) it has reached the maximum distance, so that it
%   remains at the same size.  After a Doomsday event, the signals can
%   still be observed, but the surface of last contact grows.
%   %
%   Schematic representation of two emitters, $E_1$ and $E_2$,
%   that reach each other at different times.  The time span for $E_i$
%   is $(A_i, D_i)$, for $i=\{1,2\}$.  Emitter $i$ can listen to
%   emitter $j$ between $C_{ij}$ and $B_{ij}$.
%   The type and length of causal contact in both directions depend on
%   the distance and time lag between the awakening events, the maximum
%   distance that a signal can reach and the time period in which each
%   emitter is active.
%
%
%
The initially growing communicating sphere, is shown over
space--time diagrams, where time is represented on the horizontal
direction, and space is represented in the vertical direction.
%
In the upper panel (a), the sphere is growing as the surface of first
contact has not reached the maximum distance.
%
In the middle panel (b) it has reached the maximum distance, so that
it remains at the same size.
%
After a Doomsday event (bottom panel, b), the signals can still be
observed, but the surface of last contact grows.
%
We also show in this Figure a schematic representation of two
emitters, E$_1$ and E$_2$, that reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emitter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the
distance and time lag between the awakening events, the maximum
distance that a signal can reach and the time period in which each
emitter is active.
%
In the Fig.~\ref{F_messages} we show a schematic representation of the
possible cases in which a CETI A can be in causal contact with a CETI
B which appears later.
%
The duration of the causal contact in one direction depends on several
factors, mainly $\Delta$t~A , $\Delta$t~C , and $\Delta$X.

%\begin{figure}
%   \centering
%   \includegraphics[width=0.5\textwidth]{growingsphere.pdf}
%   \caption{Schematic representation of the growing communicating
%   sphere, over the space--time diagrams, where time is represented on
%   the horizontal direction, and space is represented in the vertical
%   direction.  In (a) the sphere is
%   growing as the surface of first contact has not reached the maximum
%   distance.  In (b) it has reached the maximum distance, so that it
%   remains at the same size.  After a Doomsday event, the signals can
%   still be observed, but the surface of last contact grows }
%   \label{F_growing_sphere}
%\end{figure}
%
%  
%\begin{figure}
%   \centering
%   \includegraphics[width=0.5\textwidth]{abcd.pdf}
%   \caption{Schematic representation of two emitters, $E_1$ and $E_2$,
%   that reach each other at different times.  The time span for $E_i$
%   is $(A_i, D_i)$, for $i=\{1,2\}$.  Emitter $i$ can listen to
%   emitter $j$ between $C_{ij}$ and $B_{ij}$.
%   The type and length of causal contact in both directions depend on
%   the distance and time lag between the awakening events, the maximum
%   distance that a signal can reach and the time period in which each
%   emitter is active.}
%   \label{F_abcd}
%\end{figure}

\begin{figure}
   \centering
   \includegraphics[width=0.5\textwidth]{scheme.pdf}
   \caption{Space--time diagrams showing a schematic representation of the 
   different stages in the development of a communicating
   civilization, the growing communicating sphere (a) and the temporal
   intervals for the communication between two civilizations (b).
   %
   In the upper panel (a), the sphere the encloses the region which is
   causally connected to the emitter grows until it reaches the
   $D_{max}$ distance.  After the $D$ event, that region is limited by
   the surface of last contact.
   %
   The scheme in the lower panel (b) represents
   two communicating civilizations that have
   causal contact in both directions}   
   \label{F_scheme}
\end{figure}

The temporal structure of the process is defined by two distribution
parameters that represent the mean time interval that an intelligent
civilization can emit and receive signals, and the mean time interval
between the emergence of an intelligent comunicating civilization and
the next one.
%
The spatial structure of the simulation is given by the size and shape
of the Galactic Habitable Zone and the maximum distance a signal can
travel to be detected.
%
The parameters for the temporal distributions also determine the
spatial properties since the density of active cetis in the Galaxy
depend on these two parameters.
%
For example, a small $\tau_a$ and a large $\tau_s$ will produce a
Galaxy densely populated woth CETIs, and viceversa.
%
Also, some hypotheses must be made in order to complete the
simulation.
%
\citet{forgan_spatiotemporal_2011} argues that the times at which
different civilizations become intelligent 
follow a Gaussian distribution, and then the distribution of
inter-arrival times is an inverse exponential.
%
We chose to assume that the distribution of the times of A events is a
stationary Poisson process, and then the distribution of the times between the
appeareance of new consecutive CETIs is exponential.
%
Regarding the duration of a CETI, we propose that the distribution of
the duration of CETIs is a stationary exponential distribution.
%
\citet{balbi_impact_2018} stress the fact that it would be desirable to arrive
at a theoretical statistical distribution of the lifespan of
civilizations, preferentially on the basis of the underlying
astrophysical and biological processes.
%
\citet{maccone_lognormals_2014} argues that this distribution should be a log normal.
(EXPAND)

There is no a temporal window for the model, i.e., the simulation
starts when the process is already stable, and ends before any
galactic effect could modify or affect of fixed variables.
%
The probability for the rise of a CETI is homogeneous over the GHZ.
%
Although the Galaxy has a well known spiral structure, the assumed
sparcity of civilizations makes this assumption very reasonably.
%
Otherwise, if the distribution of civilizations is not sparce, it
could be the case that the spiral arms would host most of the
civilizations, and contacts are frecuent between closely located
civilizations.
%
In such a case we could expect that the interpretation of our results
could be considered as pesimist for the stablishment of a contact with
our planet.
%
We also limit the posibilities of life or other types of civilizations
to the hypothesis usually stated for the definition of the GHZ
\citep{dayal_habitability_2016,gonzalez_galactic_2001,lineweaver_galactic_2004,gonzalez_habitable_2005,morrison_extending_2015,haqq-misra_evolution_2019,rahvar_cosmic_2016,gobat_evolution_2016,
rahvar_cosmic_2016}.
%
\citet{vukotic_astrobiological_2012} propose probabilistic cellular
automata model to explain the astrobiological history of the Milky
Way.
%
This means that we set aside possibel civilizations that could survive
in severe conditions or unstable systems, which would prevent the
appeareance of life as we know it.

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{clccc}
\hline
   \multicolumn{2}{l}{independent variables} &min&max&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings & 4000 y & 200000 y & 50\\ 
   $\tau_{s}$ & Mean lifetime of a CETI                                 & 10000 y & 500000 y& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{3}{l}{500, 1000, 10000, 20000, 40000, 80000 lyr}  \\
\hline
   \multicolumn{4}{l}{fixed variables and assumptions} &value \\
\hline
   & statistical properties of all CETIs &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{SFC}$   & Inner radius of the GHZ     &&&20000 ly\\
   R$_{SLC}$   & Outer radius of the GHZ     &&&60000 ly\\
   t$_{max}$ & Time span of the simulation  &&&1.e6 y\\
\hline
   \multicolumn{5}{l}{discrete events}\\
\hline
   A & Awakening: a civilization starts a SETI program  &&&\\
   B & Blackout: the end of the communication chanel stops &&&\\
   C & Contact: a new causal contact is produced &&&\\
   D & Doomsday: a civilization ends its communication capabilities&&&\\
\hline

\hline
\end{tabular}
\caption{Definition of independent variables and adopted values for 
   fixed paramenters 
   that are part of the simulation.  Variable parameters define the
   spatial and temporal structure
   of the process and the maximum reach of the messages.}
\label{T_simu_hypotheses}
\end{table*}

While the aforementioned assumptions are basic conditions for most of
the stochastic processes observed in nature, we need to make stronger
assumptions related to the nature of the message.
%
The most simple assumption about the message itself, is that it
travels at light speed.
%
With this, we are considering messages sent through, for example,
electromagnetic radiation or even gravitational radiation, but we set
appart messages sent with mechanical means or physical objects, or
through some unknown technology that violates the known so far laws of
physics.
% 
For the communication of CETIs through messages sent isotropically, we
assume that the capacity to emit signals and to receive signals ocurr
at the same time.
%
This means that the ability to find another civilization develops at
the same time that the ability and intention to emit a signal in all
direction, intented to be detected by an unknown civilization.
%
Although there are several reasons to think that this could not be the
case, at large time scales it can be considered that both abilities
occurr roughly at coincident epochs.
%
Another essential assumption is that all CETIs use the same signal
power, so that there is a maximum distance out to which it can be
detected.
%
It is worth mentioning that we are considering in our experimental
setup a system composed by several emmiters and receivers across the
Galaxy, under the hypothesis proposed before.
%
In such a system we compute probabilities of a CETI making contact
with another CETI, and the results must be interpreted as for a
randomly chosen CETI, given certain conditions of homogeneityon an
ensemble of equivalent galaxies.
%
We stress the fact that we are not centered on the probabilities for
the Earth itselft, but on a generalized CETI.
%
Regarding the extent of the signals, we know that the distance from
which a signal from Earth could be detected using the current
technology is of a few hundreds light years, given that the signal was
sent to a specific direction.

\begin{figure*}
   \centering
   \includegraphics[width=0.8\textwidth]{Messages_01.pdf}
   \caption{Schematic representation of the possible cases in which a
   ceti A can be in causal contact with a ceti B which appears later.
   The duration of the causal contact in one direction depends on
   several factors, mainly $\Delta t_A$, $\Delta t_C$, and $\Delta X$.}
   \label{F_messages}
\end{figure*}

%https://setiathome.berkeley.edu/forum_thread.php?id=80585&postid=1833000#1833000]
%https://www.engadget.com/2017/05/25/listening-to-starlight-our-ongoing-search-for-alien-intelligenc/]
%[from https://en.wikipedia.org/wiki/Fermi_paradox]
%
It is straightforward to propose and implement a distribution of
maximum distances, although this would increase the model complexity
at the cost of a larger uncertainty or of another dimension on the
hypothesis space.
%
This proposal is based on a pronoid scenario (REF), and we do not
consider in this work the ocurrence of paranoid or partially paranoid
civilizations.
%
If that would be the case, the results obtained here can be taken as
upper limits.

In our simulations we also assume the simplest situations for the
growth of ETIs.
%
For example, we discard the possibility of stellar colonization
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000, maccone_mathematical_2011}
and assume that the communication aim is performed to
all directions with the same power and the same probability, i.e.,
strongly assume isotropic communication in all cases.
%
More detailed simulations could be produced considering different 
efficiency of communication or detection methods.
%
Also, a mixed paranoid/pronoid population of intelligent civilizations
could be implemented easily.
%
However, in this case we also would complicate the experimental setup,
make the results less evident.
%
The results are independent of the nature of life (organic or
artificial).
%
The lifetime of a civilization can be caused by autodestruction or by
external factors.
%}}}

\subsection{Power laws vs. exponential laws}\label{SS_PDF_shape}
%{{{

The exponential distribution of lifespan and waiting times is
justified by considering that the process of appeareance of life in
the galaxy is homogeneous and stationary.
%
This means that there is no preferred location within the GHZ for the
spontaneos appeareance of life, and that the emergence of a
civilization is independent of the existence of previous civilizations 
in the galaxy.
%            
This is equivalent to proposing a Poisson process for the emergence of
CETIs, since there is a close relation between the number of events in
time or space and the waiting time or separation, respectively.
%
That is, these are two alternative approaches to describing the same
process, a Poisson distribution for the number of events implies an
exponential distribution for their sepations, and viceversa.
%
It should be emphasized that the exponential laws used in this work
are assumed, although instead of analyzing results from a particular
parameter chosen ad--hoc, we explore the hypothesis space and analyze
the impact of the values of these parameters on the results.
%

The power law and exponential statistical distributions are among the
most common patterns found in natural phenomena.
%
For example, the distribution of the frecuency of words in many
languajes is known to follow a Zipf law (which is a power law).
%
Zipf law also describes population ranks of cities in various
countries, corporation sizes, income rankings, ranks of number of
people watching the same TV channel (Zipf's law entry on Wikipedia).
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and
floods; the sizes of meteorites or the losses caused by business
interruptions from accidents, are also well described by power laws
(Sornette 2006).
%
The power law behaviour has been observed in a variety of systems,
including for example stock market fluctuations, sizes of computer
files and word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}
%
Power laws have also been widely used in biological sciences, e.g., in
analysing connectivity patterns in metabolic networks
\citep{jeong_large_2000} and in the number of species observed per
unit area in ecology \citep{martin_origin_2006, frank_common_2009}.
% 
More examples can be found in the literature
\citep{martin_origin_2006, maccone_KLT_2010, barabasi_scale_2009,
maccone_evolution_2014, maccone_lognormals_2014,
benguigui_classificacion_2016}

\subsection{Discrete event process}
%{{{

Given a parametric model $M(\tau_a, \tau_s, D_{max})$, which includes the
functional forms of the statistical distributions and fixed values for the
hypotheses, a discrete event simulation is performed by keeping track of a set
of variables that change each time a meaningful event is produced.
%
The main variables that follow the evolution of the simulation are the positions
of stars, which are sampled randomly within the GHZ, the time of the awakening
of each CETI (A event), and the time of the vanishing of each CETI (D event).
%
The variables that can be deduced from the previous ones include the number of
CETIs in casual contact with at least another CETI at a given time, the number
of CETIs as a function of time, the number of CETIs that receive a message at
least one time, the number of CETIs that receive a message at least one time and
successfully deliver an answer and the number distribution of waiting time to
receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D)
is produced.

%%% comparar con el stochsatic cellular automata
% ver libro Solving Fermi paradox, sec. 27.3.1

\begin{figure} \centering
   %
   \includegraphics[width=0.5\textwidth]{F1.pdf}
   %
   \caption{Empirical cumulative distributions of the number contacts
   for CETIs in four different samples (M5, M6, M7, M8) with long
   range signal reach (D$_{max}$=10000).
   %
   }
   % 
\label{F_number_of_contacts} \end{figure}

\subsection{Model for CETIs}
%{{{

Since the method is based on events, the first step is to define an
architecture of events, the relationships between them and how events
trigger changes on the state of the system.
%
We consider a simulated system that represents the galactic habitable
zone (GHZ).
%
On a first approach, the GHZ is a 2-dimensional annular region.
%
The adopted values for the GHZ are 20 kly and 60 kly, for the inner
and outer radius, respectively \citep{lineweaver_galactic_2004}.
% ojo, ese paper dice 7-9 kpc, o sea tiene un limite superior mucho
% mas chico...
%
This simple model does not take into accont the variations in stellar
density given by the spiral structure.
%
Although there are several possible approaches, we chose to follow the
evolution of the system according to the following events:

\begin{enumerate}
   \item[(A)] A new CETI appears (Awakening)
   \item[(D)] An old CETI dissappears (Doomsday)
   \item[(C)] A new causal contact is stablished (Contact)
   \item[(B)] An existing causal contact is interrupted (Blackout)
\end{enumerate}

The event of type B is produced when one of the two CETIs halts in its
capability of emmiting and receiving signals.

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_1.pdf}
   \caption{Rate of CETIs that never make contact (listening) as a
   function of $\tau_a$ and $\tau_s$, for 
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{matrix_2.pdf}
   \caption{Rate of CETIs that listen at the moment of the
   awakening, as a funcion of $\tau_a$ and $\tau_s$, for
   $D_{max}$=10000 (left panel),
   $D_{max}$=40000 (middle panel), and
   $D_{max}$=80000 (right panel)}
   \label{F_C_at_A}
\end{figure*}

The system is updated each time an event is produced.
%
The time marks for each event (A and D for each CETI and C and B for each causal
contact) are stored as a result of the simulation.
%
Also, the list of active CETIs is obtained as a function of time.
%
Some fixed variables must be set in order to carry out the simulation: the size
of the Galactic Habitable Zone, the mean lifetime of a CETI, $\tau_s$, the mean
waiting time for the appeareance of another CETI, $\tau_a$, and the maximum
distance a message can be detected.

The Galactic Habitable Zone is the only one of the four parameters that is
barely known, is set at fixed values of the inner and outer radii, radial
symetry is assumed, and the with of the galactic disk is neglected compared to
the radial size.
%
The two radii are measured in light years, with the aim to maintain a single and
comprehensive unit for both space and time coordinates.

The simulation was implemented on a python 3 code, which is publically
available\footnote{https://github.com/mlares/simu\_contact}

\begin{figure}
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{waiting_s+a_dif_ylog.pdf}
   %
   \caption{Histograms of the mean waiting times for the first contact, for
   several models.
   %
   Left panel shows the histograms for several values of $\tau_a$, and
   $\tau_s$ in the range 17000-24000 yr.
   %
   Right panel shows the histograms for several values of $\tau_s$, and
   $\tau_a$ in the range 68000-100000 yr.
   %
   The shape of the histograms does not change when $\tau_a$ is
   modified, but significantly changes when $\tau_s$ is modified.
   }
   %
   \label{F_waiting_for_1C}
   %
\end{figure}

\section{Results: exploring the parameter space}\label{S_results}
%{{{

We implemented the simulation of a regular grid of models varying over
the hypothesis space, which covers 15000 different models.
%
For each model, we simulated 50 realizations with different random
seeds, in order to estimate the variance.
%
The simulated parameters, $\tau_a$, $\tau_s$ cover the ranges
4000-200000 and 10000-500000, respectively, with a partition of 50
values for each parameter.
%
For the D$_{max}$ parameter, we chose the values 500, 1000, 10000,
20000, 40000 and 80000 lyr.
%
This setup makes a total of 750000 simulations, each covering a time
range of one million years.

As a product of the simulations, several quantities are obtained.
%
Some quantities are directly derived from the discrete events, namely,
the ID of emitting CETI, the ID of emitting (receiving) CETI, the
position in the galaxy, the time of Awakening (Contact) and the time
of Doomsday  (Blackout) for each CETI.
%
This structure is always for the A-D cycle of each emitting CETI, and
repeats for each contact (C-B).
%
We can also derive the additional quantities.
%
The age of a CETI is the time elapsed from A to D.
%
%
%given time, ceti\_e is the CETI emitting signals (emiter), ceti\_r is
%the CETI receiving signals (receiver), ceti\_c is the ceti that
%listen at least another ceti (citizen) and ceti\_h is the CETI ceti
%that is lestened by at least another CETI.
%
The time span of a CETI listening another or being listened by another
can also be easily derived from the results of a realization.
%
This way we can also compute the distribution in the galaxy of cetis
that reach contact and the waiting time until the first contact or the
waiting time until the next contact.
%
Regarding the properties of the population of CETIs and its evolution,
we analyze the fraction of awaken time a ceti is listening at least
another ceti, the age of contacted CETIs at first contact, the
distribution of time to wait until next contact, the fraction of cetis
where the first contact is given at the awakening, the distribution of
the number of contacts for each CETI, the distribution of the maximum
number of contacts for each CETI, the distribution of the number of
contacts as a function of CETI age, the number of contacts as a
function of time in the galaxy, the rate of CETIs that succeed in
contact, and the distribution of distances between contacted CETIs.
%
Usefull derived quantity is the duration of two-way communication
channels the or the fraction of contacts that admit a response.
% 
We also analyze the relations between the distance to ceti vs. time of
double communication, the distance to ceti vs. age of contacted ceti,
the age of a ceti and the maximum number of contacted cetis before D,
or the lifespan of a ceti vs. max number of contacts.
%
All these quantities can be analyzed as a function of the simulation
parameters.
%
We chose eight models that are on fairly oposed regions of the
explored parameter space, and cover short/long lifetimes, short/long
waiting times for the next CETI to appear in the Galaxy and
short/large range reach of the signal, including all possible
combinations.
%
The details of each model are summarized in the
Table~\ref{T_selected_models}.

In Fig.~\ref{F_number_of_contacts} we show the Empirical cumulative
distributions of the number contacts for CETIs in four different
samples (M5, M6, M7, M8) with long range signal reach
(D$_{max}$=10000). 
%
The lifetime is more determinant than the awakening rate in increasing
the number of contacts.
%
As expected, of the four samples, the sample M7, which corresponds to
a dense awakening in the timeline and a long lifetime, is the model
that maximizes the number of contacts, reaching a maximum of order 100
contacts for a single CETI.
%
This is considerable larger than any model with a shorter lifetime,
which produce a number of contacts of at most the order of ten
contacts per CETI.
%
A couple of comments are worth to mention about this plot.
%
First, it has a logarithmic scale on the x-axis, and second it is the
cumulative, not differential, empirical distribution.

In what follows, we focus on the analysis of the duration of contacts
(Sec.~\ref{SS_multiplicity}), the waiting time for the first contact
(Sec.~\ref{SS_waiting}) and the
variations of contact posibilities as a function of the position of
the CETI in the Galaxy (Sec.~\ref{SS_location}).

\subsection{Multiplicity and duration of contacts}\label{SS_multiplicity}
%{{{

The duration of each contact for a single receiving CETI is given by the
time difference between a C event and a B event, which are direct
outputs for each simulation experiment.

\subsection{Waiting time for a first contact}\label{SS_waiting}
%{{{

In this section we analyze the distribution of the waiting times for
a first contact.
%
%Such distribution can be considered to compute the probability of
%reaching a first contact within the first XX years of a SETI project,
%assuming the methods are correct and the hypotheses of the experiment.
Such distribution can be considered to compute the probability for a
random CETI to carry out a SETI project and spend a given time until
the first contact is made.

>>>>

\documentclass[crop]{CSLB}

\usepackage{MySettings}

\input{captions}

@@

@@

A common approach to discussing the estimations of the number of
extraterrestrial intelligent civilizations (ETI) in the Galaxy has
been the use of the Drake equation \citep{gleiser_drake_2010,
prantzos_joint_2013, haqq-misra_drake_2017}.
%
The abundance of intelligent civilizations in the galaxy is a
longstanding question, which is often conceptualized as the problem of
the lack of received communication or the Fermi paradox
\citep{barlow_galactic_2012, Sotos_biotechnology_2019,
forgan_galactic_2016}.
%
There are many proposals aimed at solving this paradox, which make use
of statistical \citep{solomonides_probabilistic_2016,
vanhouten_isthere_2017, horvat_calculating_2007,
maccone_statistical_2015} or stochastic approaches
\citep{forgan_numerical_2009, bloetscher_using_2019,
glade_stochastic_2011, forgan_numerical_2010}, analytical
interpretations of the Drake equation \citep{prantzos_joint_2013,
smith_broadcasting_2009} or more speculative proposals
\citep{barlow_galactic_2013, lampton_information_2013,
conway_three_2018, forgan_galactic_2017}.
%
The uncertainties in the factors of the Drake equation make it less
prone to a formal application in order to define searching strategies
or to compute the actual number of ETIs, but is a key tool to organize
the discussion around the problem of the population of ETIs in the
Galaxy \citep{hinkel_interdisciplinary_2019}, which is noticeable
dominated by the fact that despite decades of effort not a single ETI,
if any, has been detected so far.
%
Some modifications to the original idea of the equation have tried to
imprint a stochastic nature, or to propose a probabilistic approach,
or to consider the temporal structure which is missing in the
equation.
%

@@

Temporal aspects of the distribution of communicating civilizations
and their contacts have also been explored by several authors
\citep{fogg_temporal_1987, forgan_spatiotemporal_2011,
balbi_impact_2018, balb_spatiotemporal_2018, horvat_impact_2011},
%The Impact of the Temporal Distribution of Communicating Civilizations on Their Detectability
%Temporal dispersion of the emergence of intelligence: an inter-arrival time analysis
%The spatiotemporal aspects of SETI
%
as well as efforts on considering the stochastic nature of the Drake equation
\citep{glade_stochastic_2011}.
%
The simulation approach has also benn considered
\citep{forgan_evaluating_2015, vukotic_grandeur_2016,
murante_simulating_2015, forgan_numerical_2009, forgan_galactic_2017},
although with a similar problem that consists on the large number of
parameters which are either unknown or largely uncertain.
%
%New numerical determination of habitability in the Galaxy: the SETI connection
%
%Since the factors in the Drake equation are uncertain,
%
In this work, we propose to
avoid the frequentist approach of the Drake equation and to explore a
parameter space, where instead of computing a final number, we provide
a statistical distribution that gives conditional probabilities.
%
This is an exploratory analysis that aims at providing a numerical 
tool to discuss not the several theoretical problems summarized by the
Drake equation factors, 
but the different scenarios on the basis of statistical heuristics.
%
The approach proposed here shold be considered as a compromise between
the uncertainties of the frequentist approach and the detailed recipes
required on the simulation approach.
%
We then provide a numerical framework to explore via simulations a
parameter space of unknown observables in order to discuss different
scenarios and their consecuences in terms of the probability con
making contacts.
%
The only fact that can be stated with certainty is that for the number
of years SETI projects have been working we have not received any
signal, at least
within the technical posibilities and the 
conditions stablished by SETI \citep{tarter_search_2001}.
%
This arguably astonishing result has lead to the Fermi paradox, which
states an apparent contradiction between the high estimation of the
abundance of life in the Galaxy and the lack of their detection
\citep{vanhouten_isthere_2017}.
%
The lack of detection has also ispired to a number of alternative proposals
for new SETI strategies \citep{forgan_exoplanet_2017,
balbi_impact_2018, loeb_eavesdropping_2006, maccone_KLT_2010,
tarter_advancing_2009, enriquez_breakthrough_2017, loeb_relative_2016,
maccone_SETI_2011,  lingam_relative_2019, wright_theGsearch_2015,
maccone_SETI_2013, maccone_lognormals_2014, harp_application_2018,
forgan_possibility_2013, forgan_galactic_2017, funes_searching_2019}.

@@

\input{methods}

\input{results}
